A free/open-source Kazakh-Tatar machine translation system

### 2013-04-16 paper/paper.tex
some examples for contrastive analysis of the languages, need elaboraton
some restructuring of comparitive stuff based on Ilnar's input
new bib entry, but now broken compile :(

\subsection{Morphotactic differences}
2 or 3 specific examples
%FIXME: Ilnar, I can't think of anything for this...

\subsection{Lexical and morphological differences}
2 or 3 specific examples
> present progressive
%FIXME: Ilnar, could you come up with some good examples of this?
> ... ?

\subsection{Syntactic differences}
%2 or 3 specific examples
There are a number of syntactic differences in Tatar and Kazakh, which include differences in verb valencies in equivalent translations, FIXME, and FIXME.

%FIXME: Ilnar, please check the Tatar
An example of a difference in verb valencies is with the expression corresponding to ``to like something'' in Kazakh and Tatar.  In Kazakh, the verb ‹ұна› is used, e.g. ‹оқу маған ұнайды› ``I like studying'', where the subject ``I'' in English is expressed through a dative experiencer in Kazakh and the object ``study'' in English is the grammatical subject in Kazakh.  Tatar, on the other hand, uses a verb whose arguments correspond to the arguments of ``to like'' in English, e.g. ‹мин укуны яратам› ``I like studying'', where the first person pronoun is the grammatical subject and ``studying'' is the grammatical direct object.
%(does this count as syntactic?)

@@

The morphological transducers are based on the Helsinki Finite State Toolkit \citep{hfst/2011}, a free/open-source reimplementation of the Xerox finite-state toolchain, popular in the field of morphological analysis. It implements both the \textbf{lexc} formalism for defining lexicons, and the \textbf{twol} and \textbf{xfst} formalisms for modeling morphophonological rules. It also supports other finite state transducer formalisms such as \textbf{sfst}. This toolkit has been chosen as it --- or the equivalent XFST --- has been widely used for other Turkic languages \citep{coltekin2010,altintas2001,tantug2006}, and is available under a free/open-source licence.

>>>>

\subsection{Morphological differences}
2 or 3 specific examples

> Possessive endings system (yeah!)
> Kazakh -EtIn
> Past tense system

%\subsection{Morphotactic differences}
%2 or 3 specific examples
%
%%FIXME: Ilnar, I can't think of anything for this...
%% The greetings in Tatar kept a probably older variant of /personal copula ending - question
%% partice/ sequence (Исәнмесез vs Сәлеметсіз бе?), but this doesn't seem to be a good example,
%% as it is very "special case".
%% Although, the possessive endings system is somewhat different, maybe there is something.

%\subsection{Lexical and morphological differences}
%2 or 3 specific examples
%> present progressive
%> ... ?
%
%%FIXME: Ilnar, could you come up with some good examples of this?
%% * no instrumental case in Tatar
%% * only one formal 2p pronoun in Tatar (i.e. Сез) as opposed to Kazakh "Сіз" and "Сіздер"
%% * барыпты - барган (икән), сөйлескелі келдім - сөйләшергә килдем, баратын - бара торган

\subsection{Syntactic differences}

%2 or 3 specific examples
% * Different case governance of postpositions (grep for WARNINGs in bidix)
% * музейлік экспозиция > музей экспозициясе; экспозициялық зал > экспозиция залы
% ** This works, there is a rule for that: if an /adjective + noun/ is translated with /n.attr + noun/, <px3sp> affix is added in Tatar 
% * Мен оқуды жақсы көремін > Мин укырга яратам.
% Also, see this page: http://wiki.apertium.org/wiki/Kazakh_and_Tatar/Pending_tests
There are a number of syntactic differences in Tatar and Kazakh, which include differences in verb valencies in equivalent translations, FIXME, and FIXME.

%FIXME: Ilnar, please check the Tatar
% There is "ошый" in Tatar, which works the same way as Kazakh verb.
% But I found the following in the bidix:

% <e>       <p><l>иемден<s n="v"/><s n="tv"/></l>   <r>ия<b/>бул<s n="v"/><s n="iv"/></r></p></e> <!--"" WARNING transitivity-->
% <e>       <p><l>сақтандыр<s n="v"/><s n="iv"/></l><r>кисәт<s n="v"/><s n="tv"/></r></p></e> <!--"" WARNING: IV<>TV-->
% <e>       <p><l>тұтын<s n="v"/><s n="tv"/></l>    <r>тотын<s n="v"/><s n="iv"/></r></p></e> <!--"" WARNING transitivity-->
% <e>       <p><l>тұтын<s n="v"/><s n="tv"/></l>    <r>тотын<s n="v"/><s n="iv"/></r></p></e> <!--"" WARNING transitivity-->

% There are few other WARNINGs as well.

%FIXME: use a noun in examples, not infinitive/gerund
An example of a difference in verb valencies is with the expression corresponding to ``to like something'' in Kazakh and Tatar.  In Kazakh, the verb ‹ұна› is used, e.g. ‹оқу маған ұнайды› ``I like studying'', where the subject ``I'' in English is expressed through a dative experiencer in Kazakh and the object ``study'' in English is the grammatical subject in Kazakh.  Tatar, on the other hand, uses a verb whose arguments correspond to the arguments of ``to like'' in English, e.g. ‹мин укуны яратам› ``I like studying'', where the first person pronoun is the grammatical subject and ``studying'' is the grammatical direct object.
%(does this count as syntactic?)

> Tatar infinitive (Kazakh ger+nom/acc/dat, Kazakh -GAlI vadv)

@@

The morphological transducers are based on the Helsinki Finite State Toolkit \citep{hfst/2011}, a free/open-source reimplementation of the Xerox finite-state toolchain, popular in the field of morphological analysis. It implements both the \textbf{lexc} formalism for defining lexicons, and the \textbf{twol} and \textbf{xfst} formalisms for modeling morphophonological rules. It also supports other finite state transducer formalisms such as \textbf{sfst}. This toolkit has been chosen as it --- or the equivalent XFST --- has been widely used for other Turkic languages \citep{coltekin2010,altintas2001,tantug2006,washingtonipasovtyers12,tyerswashingtonsalimzyanbattalov12}, and is available under a free/open-source licence.
